Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-04-25 joint_inf.tex
Pass through to start of second section.

We will be considering phylogenetic likelihood on unrooted phylogenetic trees.
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ distributed as $\ancestralStateColumnRV$.

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$.
Following Section 8.6 of \cite{Semple2003-em}, we take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For the binary symmetric model, a realization of $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we can start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular site split is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.
Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
For topology $\tau$, we also define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

With this split formulation, define mappings from patterns to splits as described above

@@

@@

    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\

@@

%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
%das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
%EM Perhaps we're talking about partition in different senses. I was saying that eta_i can intersect with another eta_j. Let's chat IRL on Monday.

@@

@@

@@

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
    \caption{Four taxa tree}
\end{figure}

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
%das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states. Another set of eyes might make a liar out of me, though.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_j=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
In this case, the set of most likely ancestral states is $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_j\in\ancestralSplitPartition_j(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

>>>>

We will be considering the standard phylogenetic likelihood on unrooted phylogenetic trees (cite Felsenstein book).
We assume the character alphabet $\alphabet=\{0,1\}$ and a uniform stationary distribution.
Let $\nSiteRows$ be the number of tips of the tree, and $\nAncestralStateRows = \nSiteRows-2$ the number of internal nodes.
Assume we observe $\nCols$ samples of character data (i.e.\ an alignment with $\nCols$ columns) $\fullAlignment=[\alignmentColumn_1,\ldots,\alignmentColumn_\nCols]\in\alphabet^{\nSiteRows\times\nCols}$ distributed as the random variable $\alignmentColumnRV$.
There are also corresponding unknown ancestral states $\fullAncestralStates=[\ancestralStateColumn_1,\ldots,\ancestralStateColumn_\nCols]\in\alphabet^{\nAncestralStateRows\times\nCols}$ distributed as $\ancestralStateColumnRV$.

@@

For a topology $\tau$ and branch lengths $t$ we start with the fully-observed likelihood (making the usual independence assumption between sites):
\begin{equation}
\label{eq:full_likelihood}
%EM given that the H is on the lhs of the conditioning, does it make sense to have it on the rhs of the semicolon? I don't know the convention.
L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) = \prod_{i=1}^{\nCols} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t).
\end{equation}
% In particular, we are interested in
% $$
% (\hat{\boldsymbol\xi}, \hat{\tau}, \hat{t}) = \arg\max_{\boldsymbol\xi, \tau, t} \ L_n(\mathbf{y};\boldsymbol\xi, \tau, t),
% $$
% which we call the maximum likelihood values of the parameters $(\boldsymbol\xi, \tau, t)$.
% Since the number of elements in $\boldsymbol\xi$ grows with that of the observed data $\mathbf{y}$,
The typical approach to estimate the tree $\tau$ and branch lengths $t$ involves marginalizing across all possible ancestral states
\begin{equation}
\label{eq:marginal_likelihood}
\tilde{L}_\nCols(\fullAlignment; \tau, t) = \sum_{\fullAncestralStates\in\alphabet^{\nAncestralStateRows\times\nCols}} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t)
\end{equation}
and maximizing over the topology and branch lengths to obtain
$$
(\hat{\tau}, \hat{t}) = \arg\max_{\tau, t} \  \tilde{L}_\nCols(\fullAlignment; \tau, t).
$$
% The values $\hat{\boldsymbol\xi}$ are then calculated conditional on these estimates.
% More likely in practice we fix a topology $\tau$ and use this marginalization approach to compute $(\hat{\boldsymbol\xi}, \hat{t})$.

@@

Since we have a finite character alphabet, for a given column $i$ there are a finite number of possibilities of assignments of characters to tips $\alignmentColumn_i$ or internal nodes $\ancestralStateColumn_i$; this results in a simplification of likelihood calculation (we follow Section 8.6 of \cite{Semple2003-em}).
Take the tip labels of $\tau$ to be $\{1,\ldots,\nSiteRows\}$.
For likelihood calculation under the binary symmetric model, a given $\alignmentColumn_i$ can be described as a subset of indices $\siteSplit\subseteq\siteSplitSet:=\{1,\ldots,\nSiteRows-1\}$ with equivalent characters, commonly called a ``site split.''
Indeed, we can start by including those tip labels in $\siteSplit$ that are assigned a 1 in the $\alignmentColumn_i$.
Then, because the probability of observing a particular collection of binary characters is equivalent to the probability of its complement under the binary symmetric model, we can take the complement if needed to exclude label $\nSiteRows$ from $\siteSplit$.

Define the total number of unique site splits as $\nSiteSplits=|\mathcal{P}(\siteSplitSet)|$ where $\mathcal{P}$ denotes the power set.
For topology $\tau$, we also define an ordered set of internal node labels $\{1,\ldots,\nAncestralStateRows\}$ for $\ancestralStateColumn_i$ and similarly use a subset of characters $\ancestralSplit\subseteq\ancestralSplitSet:=\{1,\ldots,\nAncestralStateRows\}$ to describe a realization of $\ancestralStateColumn_i$.
The probability of observing an ancestral state split, conditional on a site split, will not be invariant to taking its complement, and thus the need for enumerating the entire set of internal nodes.
The number of unique ancestral state splits is $\nAncestralSplits=|\mathcal{P}(\ancestralSplitSet)|$.

With this split formulation, define mappings from patterns to splits as described above:

@@

% Need to introduce \siteSplit_j and \ancestralSplit_k , no?

@@

    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)

@@

@@

Let $\xi_j$ be such a choice for each $1 \leq j \leq q$ in the following.

@@

%EM think we should say "where the site split or its complement appears."?

@@

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix. (But we should probably check with Vladimir.)
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
    \caption{Four taxon tree}
\label{fig:four-taxa-tree}
\end{figure}

@@

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
%EM Would it be worth enumerating the site splits ahead of time? It seems a little funky to be calling them j each time.
The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Thus the set of most likely ancestral states contains a single element in this case, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_j=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
In this case, the set of most likely ancestral states is $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_j\in\ancestralSplitPartition_j(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
%EM As described IRL, a little strange to be redefining (9) with an extra variable. After reading the below, I think it makes most sense to just refer to (9) indirectly but make this a stand-alone definition.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
%EM did you mean to drop the eta from the first term of this equation?
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
%EM IMHO \ell^p is so standard that it's confusing to see it in another setting. Perhaps \tilde{\ell}?
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.
